\setlength{\footskip}{8mm}

\chapter{DISCUSSION}
This is perhaps the most important parts, because it holds the insight!  It shows how smart and wise you are as a researcher.

\textbf{Write at least two pages.}

So many things you could have discussed.  This is guideline:

\section{Revisiting Research Questions}
\begin{itemize}
    \item Did you find what you expect?  Discuss in depth anything you did not expect.
    \item Did you achieve what you claim?  and what you did not yet achieve? what is your ultimate goal, and how far are you from it? 
\end{itemize}
    
\section{Comparing with Past Work}
\begin{itemize}
    \item How did your work compare with related work? consistent, anything surprising? 
\end{itemize}
    
\section{Experimental Decisions}
\begin{itemize}
    \item Why did you choose certain experimental/approaches design decisions, and how did it affect your results?
    \item You can specifically talk about certain aspects of your work 
\end{itemize}
    
\section{Design Implications}
Design implications or guidelines.   This is more tailored for HCI study.

\section{Limitations and Future Work}
\begin{itemize}
    \item Limitations: for limitations, please only describe the core limitation that you think might affect the validity of your result. Limitation such as “I conduct this study in the US but I have not tested in the UK” is not a limitation to be discussed as it is almost a common limitation in any study.  Discuss CONCRETE, REAL limitation.
    \item Future work, you only have to discuss one or two really concrete one.
\end{itemize}

